\section*{Abstract}
%%%%%%%%%%%%%%%%%%%%%%
% Was ist das Thema? Welche Methoden? % Warum relevant? Wie Evaluiert? Was für Ergebnisse/Schlüsse?
%%%%%%%%%%%%%%%%%%%%%%
Relational database systems are a powerful way of working with vast amounts of complex data. Most industries rely on big data for everyday businesses, not to mention the recent ascend of machine learning. Usually, data is stored in databases and transferred into applications to perform complex computations. By leaving the database-world, we sacrifice everything that makes databases fast: Query planer, indices, optimized I/O and more. Furthermore, we have to take the overhead of retrieving the raw-data from the database-system.

Many complex and interesting computations on big data share a common property that makes most programmers switch instantly from SQL to any other programming language: Recursion. Writing recursive algorithms as recursive queries can be very challenging since most algorithms are designed in a functional or imperative style rather than data-oriented. User Defined Functions (UDFs) were introduced to ease development of complex queries and can even handle recursion -- but recursive UDFs are incredibly slow.

With our approach we give the programmer a fully automated tool that can be used to translate a recursive function, given by an SQL-UDF that is easy to write and maintain, into a single semantically equivalent query. The translation utilizes automatic memoization and removes the overhead from recursive UDFs. A first evaluation shows its efficiency for memoizable as well as divide-and-conquere algorithms. 

This thesis provides an operational semantics to analyze a given UDF statically and extract features required to construct the translation from a set of templates. This thesis is accompanied by a first implementation in Haskell for PostgreSQL.